Accelerating Biomolecular Nuclear Magnetic Resonance Assignment with A*

### 2014-03-27 paper/tex/Paper.tex
margin fix
ALMOST DONE

\author{
Joel Venzke, Paxten Johnson, Rachel Davis, John Emmons, Katherine Roth,\\ David Mascharka, Leah Robison, Timothy Urness and Adina Kilpatrick\\
Department of Mathematics and Computer Science\\
Drake University\\
Des Moines, IA\\
email % TODO
}
\date{}

@@

\section{Introduction}
\label{sub:introduction}
Knowledge of a protein’s structure allows for further understanding of the function of a protein and alterations to its function that can lead to disease. Through the implementation of Nuclear magnetic resonance (NMR) spectroscopy, the structure of molecules, such as proteins, can be studied. However, assigning NMR spectroscopy generates large amounts of data that needs to be assigned. This process can take months to complete and is prone to error. To allow for major advances in structural biology, a faster and more accurate method needs to be developed. Our research focuses on developing an algorithm to automate the assignment process. We hope our attempts can minimize the human error and decrease the time consumption associated with manual assignment , increasing the overall effectiveness of the operation.

\subsection{Nuclear Magnetic Resonance Spectroscopy} % (fold)
\label{sub:nmr}
Several types of NMR variables can be used in the analysis of protein structures. In particular, essential information is provided by the chemical shifts of NMR-active nuclei present in proteins, including hydrogen and isotopes of carbon and nitrogen. The chemical shift is a quantifier for the deviation in the resonant frequency of a nucleus from its value in a structure-free environment, and therefore provides information on the local conformation. Determining the chemical shifts of all or most of the nuclei in a biomolecule is the first step in determining its structure. An important set of chemical shifts in a protein are those corresponding to the nuclei in the backbone of the protein polypeptide chain, including the nitrogen, attached hydrogen, and the alpha and beta carbon atoms ($C_\alpha$ and $C_{\beta}$) of each residue. These chemical shifts are measured using various three-dimensional NMR experiments, and then matched to the individual residues in the protein in a process called sequential assignment.

@@

\subsection{Data Collection} % (fold)
\label{sub:data_collection}
Determining the structure of a protein requires many highly involved steps. The first, which can take at least five days, is protein production. NMR experiments are then performed using the produced proteins in order to gather the data necessary for the assignment process. This process usually takes another one to two days for every spectrum involved. The most time consuming process is assignment of the structure. \cite{babak_alipanahi_error_2011}

\subsection{Manual Assignment} % (fold)
\label{sub:manual_assignment}
% TODO
In most cases, backbone NMR data is assigned manually. However, with hundreds of values for the chemical shift values of C(a) and C(b), this process can take anywhere from 20 days to 9 months with perfect data. To complicate matters even more, experience has shown most data sets to have missing or false peaks, requiring whomever is doing the actual assignment to either skip large portions of the protein sequence or to guess where a certain chain is located in the sequence, making this method extremely error prone. \cite{babak_alipanahi_error_2011}

@@

\subsection{Our Algorithm} % (fold)
\label{sub:algorithm}
Our algorithm begins by reading in an expected amino acid order and the NMR carbon shift data from a data file. It then changes the amino acids to their expected $C_\alpha$ and $C_{\beta}$ values and stores them in a list that represents the protein chain. Next, the carbon shift values for a single amino acid are placed in a tile. A tile contains the $C_\alpha$ and $C_{\beta}$ values for the amino acid residue $i$ and residue $i-1$. Next, missing data is accounted for. If there are fewer tiles than the number of amino acids in the protein chain, placeholder tiles are generated to make up the difference. Before the search for the best solution begins, the tiles are grouped by the types of amino acids they could represent. The groups were created using the expected $C_\alpha$ and $C_{\beta}$ values shown in Figure~\ref{fig:carbon}. Grouping is done in an attempted to accelerate the assignment process. Tiles that do not match a group are grouped with the placeholder tiles.

@@

%start search
To start the search, tiles in the placeholder group and tiles of the same amino acid group as the first amino acid in the protein chain are placed in the first position in a node. The tile is then compared to the first amino acid in the protein chain and a cost is generated. The cost is the sum of the cost of placing any previous tiles in the node, how well the current tile matched the corresponding amino acid in the protein chain and how well the tile's residue $i-1$ values match the residue $i$ values of the tile placed before it. All of the generated nodes are placed in a frontier array.

%continue assignment
The node in the frontier with the lowest cost is removed from the frontier and used to generate child nodes. To generate a child node, a tile is added to the end of the node. A child node is generated for all unplaced tiles that are either in the placeholder group or in the group matching the corresponding amino acid in the protein chain. The cost of placing the tile is then generated.  The child nodes are then added to the frontier, and the process of generating child nodes is repeated.

%goal state
Once a node has every tile placed, it is said to be in the goal state. The first node that reaches goal state is stored as the current best solution. Every node that achieves goal state is compared to the current best solution. The node with the lowest cost is then saved as the current best solution.

%solution found
A solution has been found once the current best solution's cost is lower than the cost of any nodes in the frontier. The search is then ended. The algorithm then outputs the cost of the best solution followed by the assigned carbon shift values.

@@

\subsection{Results}
\label{sub:results}
Our algorithm has been tested with NMR datasets of varying size from experiments conducted by Dr. Kilpatrick.  When compared to manually assigned data, the algorithm produces the same, optimal ordering of amino acids. Initially, we focused our research on accurate assignment at the cost of speed. Thus, at the start of our research our algorithm was only able to assign smaller datasets with a chain of ten to fifteen amino acids. Shifting our focus to optimizing the algorithm's speed while maintaining the accuracy we had managed to achieve, we were able to improve runtime. With the increased performance, we were able to correctly assign forty amino acids in less than a minute.

\subsection{Future Goals} % (fold)
\label{sub:future_goals}
In future research, we will look into maximizing speed and efficiency to the point that our algorithm can assign full amino acid sequences. In order to achieve this, we will experiment with  both parallelization (techniques) and machine learning. The latter of these two will assist in optimizing the cost calculation while the former will increase the speed of assignment. We believe these implementations will help our algorithm handle datasets with numerous missing data points, validating our results even when given incomplete data. 
% subsection future_goals (end)
\nocite{*}
% section method (end)
\bibliographystyle{amc}
\bibliography{Paper}

>>>>

\author{
Joel Venzke, Paxten Johnson, Rachel Davis, John Emmons, Katherine Roth,\\ David Mascharka, Leah Robison, Timothy Urness and Adina Kilpatrick\\
Department of Mathematics and Computer Science\\
Drake University\\
Des Moines, IA\\
joel.venzke@drake.edu
}
\date{}

@@

\section{Introduction}
\label{sub:introduction}
Knowledge of a protein's structure can explain the function of a protein and how alterations to its function can lead to disease. With the use of nuclear magnetic resonance (NMR) spectroscopy, the structure of molecules, such as proteins, can be studied. However, NMR spectroscopy generates large amounts of data that needs to be analyzed. This process can take months to complete and is prone to error. To allow for major advances in structural biology, a faster and more accurate method of analysis needs to be developed. Our research focuses on developing an algorithm to automate the assignment process. We hope our attempts can minimize the human error and decrease the time consumption associated with manual assignment, increasing the overall effectiveness of the operation.

\subsection{Nuclear Magnetic Resonance Spectroscopy} % (fold)
\label{sub:nmr}
Several types of NMR variables can be used in the analysis of proteins' structures. In particular, essential information is provided by the chemical shifts of NMR-active nuclei present in proteins, including hydrogen and isotopes of carbon and nitrogen. The chemical shift is a quantifier for the deviation in the resonant frequency of a nucleus from its value in a structure-free environment, and therefore provides information on the local conformation. Determining the chemical shifts of all or most of the nuclei in a biomolecule is the first step in determining its structure. An important set of chemical shifts in a protein are those corresponding to the nuclei in the backbone of the protein polypeptide chain, including the nitrogen, attached hydrogen, and the alpha and beta carbon atoms ($C_\alpha$ and $C_{\beta}$) of each residue. These chemical shifts are measured using various three-dimensional NMR experiments, and then matched to the individual residues in the protein in a process called sequential assignment.

@@

\subsection{Data Collection} % (fold)
\label{sub:data_collection}
Determining the structure of a protein requires many highly involved steps. The first, which can take at least five days, is protein production. NMR experiments are then performed using the produced proteins in order to gather the data necessary for the assignment process. This process usually takes another one to two days for every spectrum involved. After all the data has been gathered, the assigning of the protein's structure, the most lengthy of the steps, will then begin. \cite{babak_alipanahi_error_2011}

\subsection{Manual Assignment} % (fold)
\label{sub:manual_assignment}
% TODO
In most cases, backbone NMR data is assigned manually. However, with hundreds of chemical shift values, this process can take anywhere from 20 days to 9 months with perfect data. To complicate matters even more, experience has shown most data sets to have missing or false peaks, requiring whomever is doing the actual assignment to either skip large portions of the protein sequence or to guess where a certain chain is located in the sequence. As a result, this method is extremely error prone. \cite{babak_alipanahi_error_2011}

@@

\subsection{Algorithm Design} % (fold)
\label{sub:algorithm}
Our algorithm begins by reading in an expected amino acid sequence and the NMR chemical shift data. It then changes the amino acids to their expected $C_\alpha$ and $C_{\beta}$ values and stores them in a list that represents the protein chain. Next, the carbon shift values for a single amino acid are placed in a tile. A tile contains the $C_\alpha$ and $C_{\beta}$ values for the amino acid residue $i$ and residue $i-1$. Next, missing data is accounted for. If there are fewer tiles than the number of amino acids in the protein chain, placeholder tiles are generated to make up the difference. Before the search for the best solution begins, the tiles are grouped by the types of amino acids they could represent. The groups are created using the expected $C_\alpha$ and $C_{\beta}$ values shown in Figure~\ref{fig:carbon}. Grouping is done in an attempt to accelerate the assignment process. Tiles that do not match are grouped with the placeholder tiles.

@@

%start search
To start the search,  all the tiles matching the first amino acid in the reference protein chain are placed in the first position in a node. All placeholder tiles are also added to this position. The tile is then compared to the first amino acid in the protein chain to generate a cost. The cost is the sum of the differences generated by placing any previous tiles in the node, how well the current tile matched the corresponding amino acid in the protein chain and how well the tile's residue $i-1$ values match the residue $i$ values of the tile placed before it. All of the generated nodes are placed in an array of possible solutions, also called the frontier.

%continue assignment
The node in the frontier with the lowest cost is removed from the array and used to generate child nodes. To create a child node, a tile is added to the end of the node. A child node is generated for all unplaced tiles that are either in the placeholder group or in the group matching the corresponding amino acid in the protein chain. The cost of placing the tile is then calculated.  Next, the child nodes are added to the frontier, and the process of generating child nodes is repeated until a goal state is reached.

%goal state
Once a node has every tile placed, it is said to be in the goal state. The first node that reaches this state is stored as the current best solution. Every node that achieves goal state is compared to the current best solution. The node with the lowest cost is then replaces the previously stored best solution.

%solution found
A solution has been found once the current best solution's cost is lower than the cost of any other nodes in the frontier. The search is completed. The algorithm then outputs the cost of the best solution followed by the assigned chemical shift values.

@@

\subsection{Results}
\label{sub:results}
Our algorithm has been tested with NMR datasets of varying size from experiments conducted by Dr. Kilpatrick.  When compared to manually assigned data, the algorithm produces the same, optimal ordering of amino acids. Initially, we focused our research on accurate assignment at the cost of speed. Thus, at the start of our research, our algorithm was only able to assign smaller datasets with a chain of ten to fifteen amino acids. Shifting our focus to optimizing the algorithm's speed while maintaining the accuracy we had managed to achieve, we were able to improve runtime. With the increased performance, we were able to correctly assign forty amino acids in approximately seventy seconds.

\subsection{Future Goals} % (fold)
\label{sub:future_goals}
In future research, we will look into maximizing speed and efficiency with the goal of assigning  full amino acid sequences. In order to achieve this, we will experiment with both parallelization  and machine learning. The latter of these two will assist in optimizing the cost calculation while the former will increase the speed of assignment. We believe these implementations will help our algorithm handle datasets with numerous missing data points, validating our results even when given incomplete data.
% subsection future_goals (end)
\nocite{*}
% section method (end)
\bibliographystyle{amc}
\bibliography{Paper}
